---
title: "Data Exploration"
author: "Gaston Grant"
date: "2026-10-07"
output: html_document
---

```{r}
files <- list.files(
  "Text Files",
  pattern = "\\.txt$",
  full.names = TRUE
)

texts <- tibble::tibble(
  document = basename(files),
  text = purrr::map_chr(files, readr::read_file)
)
```




```{r}
texts$text <- stringr::str_replace_all(texts$text, "ſ", "s")



```



```{r}
library(tibble)
library(dplyr)
library(readr)
library(ggplot2)
library(quanteda)
library(quanteda.textstats)

corp <- corpus(
  texts,
  text_field = "text",
  docid_field = "document"
)

corp
```
```{r}
toks <- tokens(
  corp,
  remove_punct = TRUE,
  remove_numbers = TRUE,
  remove_symbols = TRUE
)

toks <- tokens_tolower(toks)

custom_stop <- c(
  "vnto", "haue", "doo", "hath", "bee", "ye", "thee",
  "hee", "shall", "hast", "doe", "beene", "thereof", "thus"
)

toks <- tokens_remove(
  toks,
  pattern = c(stopwords("en"), custom_stop)
)

toks
```
## Data cleaning choices

I lowercased words so capitalization would not split their counts. Following the Week 04 tutorial, I removed punctuation so forms such as “commerce.” and “commerce!” would count together. I removed numbers and symbols to reduce interference from random numbers such as page count, and since our analysis is focusing on vocabulary, numbers aren't as useful to us. I removed standard English stopwords and the tutorial’s historical stopwords to focus the comparison on content vocabulary rather than common grammatical words and to shorten the corpus in overall length. 

## Approach 1 - TF-IDF
```{r}
dfm_mat <- dfm(toks)

dfm_mat

topfeatures(dfm_mat, 25)
```

```{r}
dfm_tfidf <- dfm_tfidf(dfm_mat)

dfm_tfidf
```

```{r}

tfidf_mat <- as.matrix(dfm_tfidf)

tfidf_top_tbl <- bind_rows(
  lapply(rownames(tfidf_mat), function(doc) {
    top_terms <- sort(tfidf_mat[doc, ], decreasing = TRUE)[1:15]

    tibble(
      document = doc,
      term = names(top_terms),
      tfidf = unname(top_terms)
    )
  })
)

```

```{r}

tfidf_top_tbl %>%
  group_by(document) %>%
  summarise(
    terms = paste(term, collapse = ", "),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Distinctive Words by Document")
```
## Interpretation Questions
Do some documents share distinctive vocabulary?

Yes, documents A06785, A06789, and A06791 share words such as “moneys,” “exchange,” “realm,” and “commodities.” This suggests common disucssion of money and trade. Additionally, A32839 and A69858 share “east-india,” “company,” “charter,” and “tuns,” suggesting common discussion of the East India trade.

Are there documents whose “distinctiveness” seems driven by noise or formatting rather
than content?

Yes several of the distinctive words in document A32827 seem driven by noise or formatting such as "2dly", "pag", and "etc". Also document A06786	has words like "ll" and "ss" which seem driven by units of measure possibly. Generally, however, the documents seem driven primarly by content with a few documents exhibiting some distinction driven by noise or formatting.

## Approach 2 - Pearson

```{r}
sim_r <- textstat_simil(
  dfm_mat,
  margin = "documents",
  method = "correlation"
)

r_mat <- as.matrix(sim_r)

dim(r_mat)
round(r_mat[1:5, 1:5], 3)
```
```{r, fig.width=12, fig.height=10}
library(tidyverse)

heat_df <- as.data.frame(round(r_mat, 3)) %>%
  rownames_to_column("doc_i") %>%
  pivot_longer(
    -doc_i,
    names_to = "doc_j",
    values_to = "r"
  )

ggplot(heat_df, aes(x = doc_j, y = doc_i, fill = r)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  labs(
    title = "Pearson Correlation Between Documents",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```
```{r}
pair_indices <- which(upper.tri(r_mat), arr.ind = TRUE)

pairs <- tibble(
  document_1 = rownames(r_mat)[pair_indices[, 1]],
  document_2 = colnames(r_mat)[pair_indices[, 2]],
  correlation = r_mat[pair_indices]
)

most_similar <- pairs %>%
  arrange(desc(correlation)) %>%
  slice_head(n = 2)

least_similar <- pairs %>%
  arrange(correlation) %>%
  slice_head(n = 2)

knitr::kable(
  most_similar %>%
    mutate(correlation = sprintf("%.17g", correlation)),
  caption = "Two Most Similar Document Pairs"
)

knitr::kable(
  least_similar %>%
    mutate(correlation = sprintf("%.17g", correlation)),
  caption = "Two Least Similar Document Pairs"
)
```
## Interpretations
The most similar pair of documents are documents A32827 and A32833 which have a nearly perfect correlation score of 0.9996. The second most similar pair of documents are A32836 and A50763 with a correlation score of 0.97967.

The least similar pair of documents are A93819 and B14801__Free_Trade with a correlation score of 0.17. The second least similar pair of documents are A06791 and A93819 with a correlation score of 0.175.

According to the heatmap, the documents among the corpus seems to be fairly, soemtimes strongly, similar to one another. There are no negatively correlated documents in the entire corpus, with some even having near perfect correlation to one another. For this reason some interesting questions I would ask about the corpus would be:

1. Does the positive correlation across every pair of documents reflect a shared vocabulary of political economy, even when the texts discuss different subjects?

2. Is the entire corpus made up of documents written by only a few different authors?

3. Are all of these documents from the same time and place?


## Approach 3 - Syntactic complexity profile

Doucments selected: A06791 and A93819
```{r}
texts_df <- texts %>%
  filter(document %in% c("A06791.txt", "A93819.txt"))
```


```{r}
library(udpipe)

model_info <- udpipe_download_model(language = "english-ewt")
ud_model <- udpipe_load_model(model_info$file_model)
```


```{r}
anno_df <- texts_df %>%
  mutate(
    anno = map2(
      text, document,
      ~ udpipe_annotate(ud_model, x = .x, doc_id = .y) %>%
        as.data.frame()
    )
  ) %>%
  select(anno) %>%
  unnest(anno) %>%
  rename(document = doc_id) %>%
  select(
    document,
    paragraph_id,
    sentence_id,
    token_id,
    token,
    lemma,
    upos,
    feats,
    head_token_id,
    dep_rel
  )
anno_df
```
```{r}
syntax_df <- anno_df %>%
  mutate(
    is_word = upos != "PUNCT",

    is_clause = (upos %in% c("VERB", "AUX")) &
      str_detect(coalesce(feats, ""), "VerbForm=Fin"),

    is_dep_clause = dep_rel %in% c(
      "advcl", "ccomp", "xcomp", "acl", "acl:relcl"
    ),

    is_coord = dep_rel %in% c("conj", "cc"),

    is_complex_nominal = dep_rel %in% c(
      "amod", "nmod", "compound", "appos"
    )
  )
syntax_df
```

```{r}
sentence_df <- syntax_df %>%
  filter(is_word) %>%
  group_by(document, sentence_id) %>%
  summarise(
    words = n(),
    clauses = sum(is_clause),
    dep_clauses = sum(is_dep_clause),
    coordination = sum(is_coord),
    complex_nominals = sum(is_complex_nominal),
    .groups = "drop"
  )
sentence_df
```

```{r}
all_measures <- sentence_df %>%
  group_by(document) %>%
  summarise(
    MLS = mean(words),
    C_per_S = sum(clauses) / n(),
    DC_per_C = sum(dep_clauses) / pmax(sum(clauses), 1),
    DC_per_S = sum(dep_clauses) / n(),
    Coord_per_C = sum(coordination) / pmax(sum(clauses), 1),
    Coord_per_S = sum(coordination) / n(),
    CN_per_C = sum(complex_nominals) / pmax(sum(clauses), 1),
    CN_per_S = sum(complex_nominals) / n(),
    .groups = "drop"
  )

all_measures %>%
  mutate(across(where(is.numeric), ~ sprintf("%.17g", .x))) %>%
  knitr::kable(caption = "Syntactic Complexity Measures")
```
```{r}
texts_df %>%
  mutate(sentences = stringr::str_split(text, "(?<=[.!?])\\s+")) %>%
  tidyr::unnest(sentences) %>%
  filter(
    (document == "A06791.txt" &
       stringr::str_detect(sentences, fixed("The like would happen"))) |
    (document == "A93819.txt" &
       stringr::str_detect(sentences, fixed("have Indicted the Defendant")))
  ) %>%
  select(document, sentences) %>%
  knitr::kable()
```
## Syntactic Difference
A06791 more often builds sentences by adding conditions and explanations to a main statement. A93819 more often joins actions or items together and adds details describing people, places, and things. A06791 uses more dependent clauses, while A93819 uses more coordination and noun modifiers. In A06791, ‘if the money were embased by allay’ adds a condition to the main statement. In A93819, ‘for stopping,’ ‘for planking,’ and ‘for erecting’ join three actions into a list. These examples show the difference between adding dependent clauses and joining actions together.

## Interpretation
How do the two texts differ in syntactic complexity?

The sentences in both texts are about the same length, but they are strung together differently. A06791 uses more clauses to add conditions and explanations. A93819 joins more actions or items together and adds more details to nouns. The syntactic complexity of both documents is similar, but their depth of complexity is achieved in different ways.


Do these differences align with or complicate your earlier lexical findings?

These differences both align with and complicate the earlier lexical findings. These two texts were among the least similar in vocabulary across the corpus, and they also build their sentences differently. However, both use long, complex sentences, showing that they share some similarities in sentence structure despite their differences in vocabulary. 


What kinds of rhetorical or stylistic practices might these syntactic patterns reflect?

A06791’s use of conditions and explanations may reflect an effort to persuade readers by laying out a logical argument about money and trade. A93819’s list of actions and detailed descriptions may reflect the writer's need to clearly identify the accusations, people, and places involved, such as in a courtroom or trial. Both use long sentences, but for different purposes.

## Synthesis Question

Do differences in subject matter lead to differences in syntactic complexity?

## Discussion of Evidence

Based on our analysis on the distinctive words in A06791 and A93819, the two documents differed in vocabulary greatly. A06791 seemed more focused on trade and commerce having distinctive words such as "moneys", "exchange", "silver", and "shillings". While A93819 seems to be about a legal dispute having used words such as "defendant" and “indicted”. Among the entire corpus, every pair of documents had positive Pearson r correlation, however, the strength varied. A06791 and A93819 were among the least similar pairs. Their word patterns differed more than those of most other pairs, even though they still shared some vocabulary. Both documents had similar sentence length at about 43 words on average per sentence, however, A06791 had more dependent clauses per sentence, while A93819 had more coordination and noun modifiers. Both documents use complex sentences, but the nature in which sentences are built in each document differs. The analysis of these two documents suggests that differences in subject matter may relate to how sentences are built, without one document being consistently more syntactically complex than the other. Further investigation could examine whether texts about more serious subjects use more complex sentences.